---
title: "Exploring the Vice data"
output: html_document
---

## Packages

```{r}
library(here)
library(dplyr)
library(readr)
library(ggplot2)
```

## Read in the data

Columns in common:

* Date
* Number of Subjects (It's always 1 in vice subject dataset).
* Fatal
* Subject Armed
* Notes
* Subject Race
* Subject Gender
* Subject Age
* Nature of Stop
* Number of Shots
* Number of Officers
* Officer Race
* Officer Gender
* Department
* Full Narrative
* City

Columns unique to vice incidents:

* year, month, day: turns the date into 3 columns.
* fdate: turns dates which are missing month and/or day into a fixed date.
* b, l, w, and a: Total number of subjects in each racial category.
* tr: Total number of racial categories identified. Mostly agrees with Number of Subjects, but it will differ if racial categories are missing.
* AvgAge: The average age of the subjects.
* VA: Binary variable indicating if any subject was armed or not.
* F: Binary variable indicating if any subject died.
* nshots: Total number of shots. Turns NumberOfShots into a continuous, numeric variable.
* sit: Categorizes Nature of Stop into a more manageable set of categories.
* other: Was a weapon other than a knife/gun/replica weapon used? Taken from notes and narrative.
* knife: Was a knife used? Taken from notes and narrative.
* replica: Was a replica weapon used? Taken from notes and narrative.
* weapon: What weapon was used? Taken from notes and narrative.

```{r}
vice_incident <- read_csv(here("data_cleaning", "data", "vice-incident.csv"))
vice_subject <- read_csv(here("data_cleaning", "data", "vice-subject.csv"))
```

