---
title: "Exploring the Washington Post Data"
output: pdf_document
---

## Packages

```{r, message = F, warning = F}
library(here)
library(readr)
library(dplyr)
library(purrr)
library(tidyr)
library(ggplot2)
library(lubridate)
```

## Read in data

```{r}
washpo <-
    read_csv(here("data_cleaning", "data", "washpo.csv")) %>%
    select(-id, -name, -longitude, -latitude, -is_geocoding_exact, -city) %>%
    mutate(year = year(date),
           month = month(date),
           day = day(date)) %>%
    select(-date)
```

## Distribution of victim age

```{r}
washpo %>%
    select(age) %>%
    pivot_longer(everything()) %>%
    ggplot(aes(x = value)) +
    geom_histogram(bins = 25) +
    facet_wrap(~name, scales = "free_x") +
    theme_bw()

summary(washpo$age)
```

## Distribution of victim age by race

* Black American and Hispanic victims seem to be younger.

```{r}
washpo %>%
    select(age, race) %>%
    filter(race %in% c("W", "B", "H")) %>%
    ggplot(aes(x = age)) +
    geom_histogram(bins = 25) +
    facet_wrap(~race, scales = "free", ncol = 1) +
    theme_bw()

map(c("W", "B", "H"),
    function(df, raceCat) {
        df <- df %>% filter(race == raceCat)
        print(raceCat)
        summary(df$age)},
    df = washpo)
```

## Bar charts for categorical outcomes (not state, armed, or date)

```{r}
washpo %>%
    select(c(!where(is.numeric), -armed, -state, -day, -month, -year)) %>%
    pivot_longer(everything(),
                 values_transform = list(value = as.character)) %>%
    ggplot(aes(x = value)) +
    geom_bar() +
    facet_wrap(~name, scales = "free") +
    theme_bw()
```

## Bar charts for categorical outcomes by race (not state, armed, or date)

* Slightly more instances in which body cameras were used for Black and Hispanic victims vs. White victims.
* White victims slightly more likely to be fleeing vs. Black and Hispanic victims.
* White victims slightly, slightly, less likely to be male vs. Black and Hispanic victims.
* Manner of death is universally the same.
* White victims moderately more likely to have a mental health history or signs of mental illness vs. Black and Hispanic victims.
* Similar levels of threat presented by race however the types of threat vary slightly across race.

```{r}
categoriesRace <- 
    washpo %>%
    select(c(!where(is.numeric), -armed, -state, -day, -month, -year)) %>%
    pivot_longer(-race,
                 values_transform = list(value = as.character)) %>%
    filter(race %in% c("W", "H", "B")) %>%
    count(race, name, value) %>%
    group_by(race, name) %>%
    mutate(prcnt = n / sum(n)) %>%
    ungroup()

map(unique(categoriesRace$name),
    function(df, column) {
        df %>%
            filter(name == column) %>%
            ggplot(aes(x = value, y = prcnt)) +
            geom_bar(stat = "identity") +
            facet_wrap(~name+race, scales = "free_x") +
            theme_bw()},
    df = categoriesRace)
```

## Distribution of victims by state

```{r}
washpo %>%
    select(state) %>%
    group_by(state) %>%
    count(state) %>%
    ggplot(aes(x = reorder(state, n), n)) +
    geom_bar(stat = "identity") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 290, hjust = 0))
```

## Distribution of victims by state by race

* Within each racial category, what proportion of killings took place in each state? E.g. of all Black Americans killed by the police, 10% were killed in CA. Of all White Americans killed by the police, 8& were killed in California. Of all Hispanics killed by the police, over 30% were killed in CA.
* Hard to discern any immediate trends. The biggest thing which stands out is the fact that Hispanics are overwhelmingly killed in only a small proportion of stats. Possibly tied to where Hispanic populations are largest.

```{r}
washpo %>%
    select(state, race) %>%
    filter(race %in% c("W", "B", "H")) %>%
    count(state, race) %>%
    group_by(race) %>%
    mutate(prcnt = n / sum(n)) %>%
    ungroup() %>%
    ggplot(aes(x = reorder(state, prcnt), y = prcnt)) +
    geom_bar(stat = "identity") +
    facet_wrap(~race, scales = "free_y", ncol = 1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 290, hjust = 0))
```

* Within each state, what proportion of killings were for each race? E.g. of all Americans killed by the police in Alabama, 25% were Black, 1% were Hispanic, and 70% were White.
* Hard to discern any immediate trends. I need population totals to get a sense of the disproportionality.

```{r}
washpo %>%
    select(state, race) %>%
    filter(race %in% c("W", "B", "H")) %>%
    count(state, race) %>%
    group_by(state) %>%
    mutate(prcnt = n / sum(n)) %>%
    ungroup() %>%
    ggplot(aes(x = reorder(state, prcnt), y = prcnt)) +
    geom_bar(stat = "identity") +
    facet_wrap(~race, scales = "free_y", ncol = 1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 290, hjust = 0))
```

## Number of killings by day, month, year

```{r}
washpo %>%
    select(day, month, year) %>%
    pivot_longer(everything()) %>%
    ggplot(aes(x = value)) +
    geom_bar() +
    facet_wrap(~name, scales = "free") +
    theme_bw()
```

## Number of killings by day, month, year and by race

I wasn't expecting there to be any disparity by date and race, but it sure would've been interesting if there was disparity.

```{r}
washpo %>%
    select(day, month, year, race) %>%
    pivot_longer(-race) %>%
    filter(race %in% c("W", "B", "H")) %>%
    ggplot(aes(x = value)) +
    geom_bar() +
    facet_wrap(~name+race, scales = "free") +
    theme_bw()
```

## Distribution of victims by if they were aremed or not

```{r}
armedNewDf <-
    washpo %>%
    select(armed) %>%
    count(armed) %>%
    mutate(prcnt = n / sum(n),
           armedNew = if_else(prcnt < 0.01, "other", armed))

armedNewDf %>%
    group_by(armedNew) %>%
    summarise(n = sum(n)) %>%
    ungroup() %>%
    ggplot(aes(x = reorder(armedNew, n), y = n)) +
    geom_bar(stat = "identity") +
    theme_bw() +
    coord_flip()
```

## Distribution of victims by if they were aremed or not and by race

No big differences in types of weapons used it seems. Although Black-Americans are slightly, slightly more likely to be unarmed.

```{r}
washpo %>%
    select(armed, race) %>%
    filter(race %in% c("W", "B", "H")) %>%
    inner_join(select(armedNewDf, armed, armedNew), by = "armed") %>%
    count(armedNew, race) %>%
    group_by(race) %>%
    mutate(prcnt = n / sum(n)) %>%
    ungroup() %>%
    ggplot(aes(x = reorder(armedNew, prcnt), y = prcnt)) +
    geom_bar(stat = "identity") +
    facet_wrap(~race) +
    theme_bw() +
    coord_flip()
```
